Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-25 sections/eval.tex
Update on Overleaf.

\subsubsection{Finch Interpreter}

\textbf{Lowering}

\textbf{Heuristic Optimization}

\subsubsection{Finch Interpreter}

Find an example where fusing the python interface gives a big speedup over non-fused kernels.

@@

\begin{figure}
	\includegraphics[width=\linewidth]{elementwise_speedup_over_duckdb.png}
    \caption{Performance of Finch on elementwise (A+B)*C kernel.}
\end{figure}

@@

>>>>

\subsubsection{Standardizing \& Heuristic Optimization}
Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves the above questions about loop ordering and output formatting. This standard form has a few syntactic requirements, but the semantically important requirements are 1) all inputs (i.e. tables and alias operators) in a query's RHS must conform to a common ordering of the indices 2) the outermost operator of each query's RHS must be a reformat 3) the expression within the reformat must be a pointwise expression, optionally wrapped in an aggregate operator. The former allows the interpreter to identify the loop order for each kernel. The second determines the output format for each intermediate. The last one guarantees that the innermost expression can be computed as a single kernel.

During the standardization process, a concordization pass is performed which examines each query in order and selects a loop order based on a heuristic which loops over intersecting variables first. Then, each query is examined again and a transposition query is inserted for each input which doesn't match the loop order and then the input is replaced with an alias. At this point, the first semantic requirement is satisfied. Next, a formatting pass is performed which examines each query and selects a level format for each output index based on the formats of the inputs and whether random writes are required. The procedure for this is a simple set of rules which attempts to aggressively preserve the structure present in the input tensors.

\subsubsection{Finch Interpreter} 
Once the Finch Logic program has been converted to the standard form, the Finch Interpreter can execute each query, in order, through a straightforward lowering process. First, the output format is identified by unpacking the outer $\finchreformat$ statement. Next, the inner expression of the is unpacked to identify the aggregate operator and to convert the pointwise expression into a Finch expression. At this point, any aliases to the result of previous queries are replaced with an access to the actual tensor. Lastly, the concordant loop order is identified and instantiated. At this point, the lowered query can be compiled and executed with the Finch compiler, and the result is assigned to $name$ in the plan's scope before proceeding to the next query in the plan.

\subsubsection{The Benefits of Kernel Fusion}

@@

\begin{figure}
	\includegraphics[width=\linewidth]{elementwise_speedup_over_duckdb.png}
    \caption{Performance of Finch on elementwise (A+B)*C kernel as the sparsity of C varies.}
\end{figure}

@@

\willow{Note: I may want to explain format inference somewhere in here, but I'll have to get to it a little later, perhaps after the paper deadline}

### 2024-03-25 sections/intro.tex
Update on Overleaf.

  \begin{table}[h!]
  \centering
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{StructTensor} & \textbf{Finch} \\
  \hline
  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  Padded                   & \checkmark &            &            &            &            & \checkmark \\
  One Sparse               &            & \checkmark &            & \checkmark &            &\checkmark \\
  Sparse                   &            & \checkmark &            &            &            &\checkmark \\
  Run-length               &            &            &            &            &            & \checkmark \\
  Symmetric                &            &            &            &            & \checkmark & \checkmark \\
  Regular Sparse Blocks    &            & \checkmark &            &            &            & \checkmark \\
  Irregular Sparse Blocks  &            &            &            &            &            &\checkmark \\
  Ragged                   &            &            & \checkmark &            &            & \checkmark \\
  \end{tabular}
  \caption{Support for various data structures across tools. Finch supports \textbf{both} complex programs and complex data structures.}
  \label{tab:data_structures}
\end{table}
\teo{ New paragraph:
Note: This first sentence is still problematic; we need to say something other than tightly integrated so that we can be different from that. But I am not sure what.
%
Many of these special purpose compilers represent a program's control flow via a specialized data structure that is tightly integrated with the class of supported data structures.
%
For example, TACO's merge lattices allows it to represent iterations over subsets of rectangular grids where non-zero values are produced whereas the polyhedral model allows various compilers to represent dense computations on irregular regions~\cite{kjolstad_tensor_2017, grosser2012polly}.
%
Taichi represents general control flow with specialized instruction for accessing a single sparse structure drawn from a particular class, which is why it supports the most control flow for a single sparse structure~\cite{hu_taichi_2019}.
%
Furthermore, the limits of a DSL can be traced to the tight coupling of a program's control flow with the data structure: many new TACO features required modifying algorithms that worked with TACO's merge lattices and Tiramisu can express programs that Halide can't because Tiramisu uses a polyhedral model as opposed to a model built on iterating over regular grids~\cite{kjolstad_tensor_2017, henry_compilation_2021, senanayake2020sparse, baghdadi2019tiramisu}.
%
In this work, we will choose find a new representation of a program's data structures and control flows that covers a wide range of programs.
%
However, before introducing this, we must discuss performance: if we want to also produce efficient code for structured data, we face two challenges.
}

>>>>

  \begin{table}[h!]
  \centering
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{StructTensor} & \textbf{Finch} \\
  \hline
  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  Padded                   & \checkmark &            &            &            &            & \checkmark \\
  One Sparse               &            & \checkmark &            & \checkmark &            &\checkmark \\
  Sparse                   &            & \checkmark &            &            &            &\checkmark \\
  Run-length               &            &            &            &            &            & \checkmark \\
  Symmetric                &            &            &            &            & \checkmark & \checkmark \\
  Regular Sparse Blocks    &            & \checkmark &            &            &            & \checkmark \\
  Irregular Sparse Blocks  &            &            &            &            &            &\checkmark \\
  Ragged                   &            &            & \checkmark &            &            & \checkmark \\
  \end{tabular}
  \caption{Support for various data structures across tools. Finch supports \textbf{both} complex programs and complex data structures.}
  \label{tab:data_structures}
\end{table}
\teo{ New paragraph:
%
Many of these special purpose compilers represent a program's control flow via a specialized data structure that is tightly integrated with a narrow class of supported data structures.
%
For example, TACO's merge lattices allows it to represent iterations over subsets of rectangular grids where non-zero values are produced whereas the polyhedral model allows various compilers to represent dense computations on irregular regions~\cite{kjolstad_tensor_2017, grosser2012polly}.
%
Taichi represents general control flow with specialized instruction for accessing a single sparse structure drawn from a particular class, which is why it supports the most control flow for a single sparse structure~\cite{hu_taichi_2019}.
%
Furthermore, the limits and advantages of a DSL can be traced to the specific coupling of a program's control flow with the data structure: many new TACO features required modifying algorithms that worked with TACO's merge lattices and Tiramisu can express programs that Halide can't because Tiramisu uses a polyhedral model as opposed to a model built on iterating over regular grids~\cite{kjolstad_tensor_2017, henry_compilation_2021, senanayake2020sparse, baghdadi2019tiramisu}.
%
In this work, we will choose find a new representation of a program's data structures and control flows that covers a wide range of programs.
%
However, before introducing this, we must discuss performance: if we want to also produce efficient code for structured data, we face two challenges.
}
